Word Identification Under Multimodal Uncertainty

### 2017-01-31 write-up/main.tex

\title{Word identification under multimodal Uncertainty}

@@

Language requires the ability to process arbitrary multimodal input. It also requires the ability to infer categories under uncertainty. In this paper, we introduce a task that allows us to examine how adults identify word categories under uncertainty from the auditory and the visual modality. We propose a Bayesian model of the task which provides an optimal baseline. The predictions are tested in two experiments where category recognition is made under two kinds of uncertainty: ambiguity and noise. In both cases, human data shows many patterns of optimal behaviour, and the baseline explains most of the variance. However, when one modality is noisy, we found evidence for a systematic preference for the non-noisy modality.

@@

Language uses symbols expressed in one modality (e.g., the auditory modality in the case of speech) to communicate about things in the world, which we perceive through  various sensory modalities (e.g., the visual modality in the case of concrete objects). Therefore, understanding language requires the ability to process arbitrary multisensory input. This processing, however, is only a first step. When a speaker utters a word and points to an object, it is not enough for the observer to be able to encode these precise sound and object tokens. They should, on top of that, be able to identify the underlying categories (such as the phonemes, the concept, and the word category). This challenge is quite obvious when a person has to learn these categories (e.g., a baby who starts learning language). But a challenge is present even when the person already knows the categories, but has to recognize them. This is because individual tokens are often ambiguous or/and noisy\cite<e.g.,>{hillenbrand1995}. They do not indicate with absolute certainty their category membership, they do so only in probabilistic terms. For example, the speaker may make a mispronunciation, or their utterance may get distorted because of ambient noise. Similarly, the identity of the object that is being referred to may be uncertain when there exist other surrounding and/or similar looking objects, or when the observer's view is partially obstructed.

It follows that, in order to identify the speaker's intended categories, an observer has to be able to reason under a great deal of uncertainty. To characterize this ability, it has been suggested that people encode tokens with their probability distributions, base their decisions on these probabilities, and adjust their behavior when these probabilities change \cite<e.g.,>{clayard08}. However, previous work in this line of research has largely focused on the unimodal case, that is, when categories can be defined along one modality, such as the phonemes \cite<e.g.,>{feldman2009}. A few studies did explore some aspects of cross-modal processing in this framework \cite<e.g.,>{bejjanki2011, kleinschmidt2015}, but focused on the specific case of phoneme recognition from speech and lip movement, wherein people assume information from both modalities to be tightly correlated.

In the present work, we study the case of word category recognition, where a word is an \textit{arbitrary} mapping between a sound category and a visual category. We test people on their ability to encode audio-visual tokens and use them to recognize the underlying word category. Crucially, this recognition is made under uncertainty from both the auditory category (the phoneme) and the visual category (the concept being referred to). Our goal is to explore how adults deal with such multimodal ambiguity in a categorical task. To this end, we propose a model that combines uncertainty in an optimal way, and we use it as a baseline for human data. More precisely, we examine the extent to which human data conform to, or deviate from the baseline, and whether the deviation involves a systematic preference for a particular modality?

The paper is organized as follows. First, we introduce our audio-visual word recognition task. Second, we present the model and explain how it generate an optimal baseline. Then we present the behavioral experiments where we test category recognition under audio-visual uncertainty. In experiment 1, audio-visual tokens are clear but their category membership is ambiguous.  In experiment 2, input to one modality is ambiguous, and input to the other is ambiguous as well as noisy. We analyse and discuss the results of the experiments in the light of the model's predictions.

\section{The audio-visual word recognition task}

@@

In each trial, participants are presented with an audio-visual target (the prototype of the target category), immediately followed by an audio-visual test. 
The test may differ from the target in both the auditory and the visual components. The auditory component is sampled from a continuum along the second formant (F2) linking the words /aba/ and /ada/. The visual component is sampled from a morph that links the prototypical pictures of the dog and the cat.  After these two presentations, participants are instructed to press `same' or `different'.

This task is similar to the task introduced by \citeA{sloutsky2003} and used in subsequent research to probe audio-visual encoding. However, unlike this previous line of research, here participants are not asked whether the two audio-visual presentations are identical, instead, they are asked to press `same' if they think the second item (the test) belonged to the same word category as the prototype (e.g.,  dog-/aba/), even if there is a slight sound mispronunciation,  a slight difference in the object’s shape, or a slight difference in both the pronunciation and the shape. They are instructed to press `different' if they think that the test was, rather, an audio-visual instance to the other word category (cat-/ada/).

As shown in figure \ref{fig:task}, besides the audio-visual trials where both the sounds and the pictures are presented, there are also unimodal trials where the pictures are hidden (auditory-only) or where the sounds are muted (visual-only). These unimodal trials are crucial since they provide us with the probability distributions of the auditory and visual categories over their tokens. In fact, the goal is to examine whether participants' responses on the bimodal trial are optimal combination of the probabilities derived from their responses on the unimodal trials.

Next, we explain how we modeled this task in a Bayesian framework.

@@

\section{What would an ideal recognizer do?}

The unimodal conditions (auditory-only or visual-only)  are used to characterize the probability distributions of the auditory and the visual categories.  We assume, for the sake of simplicity, that the probability distributions are normally distributed:

$$ a | A \sim  N(\mu_A, \sigma^2_A) $$
$$ v | V \sim  N(\mu_V, \sigma^2_V) $$

In each modality, we have two categories: /ada/ ($A=1$) and /aba/ ($A=2$) in the auditory dimension, and \textit{cat} ($V=1$) and \textit{dog} ($V=2$) in the visual dimension (the identity of the categories was randomized in the experimental setup).

In the bimodal condition, participants see word tokens made of audio-visual input, and have to make a categorical decision. We represent word tokens as vectors in the audio-visual space $\mathbf{w}=(a,v)$.
A word category $W$ is defined as the joint distribution of the auditory and visual categories. It can be characterized with a bivariate normal distribution in the audio-visual space:
$$ \mathbf{w} | W \sim  N(M_W, \Sigma_W) $$
\begin{figure}[tp]
\centering
\includegraphics[width=0.4\textwidth]{MyTask.png}
\caption{Model of the task: a word category is defined as the joint bivariate distribution of an auditory category (bottom) and a visual semantic category (left). Upon the presentation of a word token $w$, participants guess whether it is sampled from the word category $W1$ or from $W2$. Decision threshold is where the guess probability is 0.5. (Plotted distributions are for illustration only)}
\label{fig:space}
\end{figure}
We have two word categories: dog-/aba/ ($W=1$) and cat-/ada/ ($W=2$). Participants can be understood as choosing one of these two word categories (figure \ref{fig:space}). For an ideal ``recognizer'', the probability of choosing category 2 when presented with an audio-visual instance $\mathbf{w}=(a,v)$ is the posterior probability of this category:
\begin{equation}
\small
p(W=2 | \mathbf{w})=\frac{p(\mathbf{w}|W=2)p(W=2)}{p(\mathbf{w}|W=2)p(W=2)+p(\mathbf{w}|W=1)p(W=1)}
\end{equation}
We make the assumption that, given a particular word category, the auditory and visual tokens are independently generated:
\begin{equation}
p(\mathbf{w} | W) = p(a,v| W) = p(a| W)p(v| W)
\end{equation}

Under this assumption, the posterior probability reduces to:
\begin{equation}
p(W=2 | \mathbf{w})=\frac{1}{1+(1+\epsilon)\exp(\beta_0+\beta_aa+\beta_vv)}
\label{eq:pred}
\end{equation}

@@

and $1+\epsilon=\frac{p(W=1)}{p(W=2)}$ is the proportion of the prior probabilities. If the identity of word categories is randomized, and if $W1$ is the target, then $\epsilon$ measures a response bias to `same' if $\epsilon > 0 $, and a response bias to `different' if $\epsilon < 0 $.

@@

Next, we present the behavioral experiments. Our goal is twofold: 1) examine the extent to which human responses correspond or deviate from the predictions of the optimal baseline. 2) if there is a deviation from the baseline, examine whether this deviation involves a systematic preference for a particular modality.

\section{experiments}

In experiment 1, we test the predictions of the model in the case where uncertainty is due to ambiguity in terms of category membership.

\subsection{experiment 1}

\subsubsection{Participants}
We planned to run 100 participants recruited from Amazon Mechanical Turk. Only participants with US IP addresses and a task approval rate above 85\% were allowed to participate. They were paid at an hourly rate of \$6/hour.  Data
were excluded if participants completed the task more than
once (2 participants). Moreover, as specified in the preregistration (\url{https://osf.io/h7mzp/}), participants were excluded if they had less than 50\% accurate responses on the unambiguous training trials (6), and if they reported having experienced a technical problem of any sort during the online experiment (14). The final sample consisted of 76 participants.

\subsubsection{Stimuli}
For the auditory stimuli, we used the continuum introduced in \citeA{vroomen2004}. It is a 9-point /aba/--/ada/ speech continuum which was created by varying the frequency of the second (F2) formant in equal steps. We selected 5 equally spaced points from the original continuum by keeping the end-points (prototypes) 1 and 9, as well as the three points: 3, 5, and 7 along the continuum.

For the visual stimuli, we used a morph continuum introduced in \citeA{freedman2001}. In the original stimuli, a continuous set of images was generated from three cat prototypes and
three dog prototypes. We selected a morph along the dog and cat prototypes that seemed most natural. Moreover, from the original 14 points, we selected 5 points as follows: we kept the item that seemed most ambiguous (point 8), the 2 preceding points (i.e., 7 and 6) and the 2 following points (i.e., 8 and 9). The 6 and 9 points along the morph were quite distinguishable, and we took them to be our prototypes (figure \ref{fig:morph}).

\begin{figure}[tp]
\centering
\includegraphics[width=0.4\textwidth]{morph.png}
\caption{Visual stimuli}
\label{fig:morph}
\end{figure}

\subsubsection{Design and Procedure}
We told participants that an alien was naming two objects: a dog, called /aba/ in the alien language, and a cat called /ada/. In each trial, we presented the first object (the target) on the left side of the screen simultaneously with the corresponding sound. The target was always the same (e.g., dog-/aba/). The second sound-object pair (the test) followed after half a second. It showed up on the right side of the screen and was supposed to be either another dog-/aba/, or a cat-/ada/. For both the target and the test, we displayed the object when the corresponding sound started playing, and we made it disappear when the sound stoped (for a total of around 800ms). For each participant, we randomized the sound-object mapping as well as the identity of the target. Finally, we instructed participants to press S (for same) if they thought the alien was naming another dog-/aba/, and D (for different) if they thought the alien was naming a cat-/ada/.

The first part of the experiment was a training to familiarize the participants with the task and the stimuli. In this training phase, we used only the prototype pictures and the prototype sounds, and each trial had a correct answer: the test item was either exactly the same as the target, or the prototype of the other word category. There were a total of 12 trials, half same, and half different. The second part was the real task. In this part, we told participants that, like in the training, the first sound-object pair would be the same. However, in the second pair, both the sound and the picture could be ambiguous. We told them that each trial was different. Finally, we encouraged participants to base their answers on both the sounds and the pictures in the bimodal condition. Each test item was made of a sound sampled from the /aba/-/ada/ continuum, and the picture form the dog-cat morph. There was a total of 25 possible combinations in the bimodal condition, and 5 in each of the unimodal conditions. Each participants answered twice to each possible trial in each condition for a total of 70 trials/participants. For each participant,  we randomized the order of conditions (auditory-only, visual-only, and bimodal). We designed the experiment to take between 10 to 15 minutes to complete.
\begin{figure}
\centering
\includegraphics[width=0.4\textwidth]{unimodal.png}
\caption{Average human responses in the auditory-only condition (left), and visual-only condition (right). Error bars are 95\% confidence intervals. Solid lines represent the unimodal fits. }
\label{fig:unimodal}
\end{figure}

\subsubsection{Results}

\begin{figure}[h]
\centering
\includegraphics[width=0.4\textwidth]{correlation.png}
\caption{Human responses vs. the baseline predictions. Error bars are 95\% confidence intervals. In experiment 1, the baseline and the fit explained 89\%, and 94\% of total variance, respectively. In experiment 2, the baseline and the fit explained 91\%, and 97\% of total variance, respectively}
\label{fig:corr}
\end{figure}

\begin{figure*}
\centering
\includegraphics[width=0.83\textwidth]{Exp.png}
\caption{Proportion of `different' as a function of the auditory distance. Solid lines represent average human responses (left), predictions of the baseline (middle), and the bimodal fit (left). Dotted lines represent average human responses in the unimodal conditions. Colors represent values in the visual continuum. }
\label{fig:Exp}
\end{figure*}

\textit{Unimodal conditions:} Let's take the auditory-only case as an example, participants can be understood as choosing one of two auditory categories. For an ideal recognizer, the probability of choosing category 2 (that is, to answer `different') when presented with an audio instance $a$ is the posterior probability of this category $p(A=2|a)$. If we assume that both categories have equal variances, the posterior probability reduces to:
\begin{equation}
p(A=2 | a)=\frac{1}{1+(1+\epsilon_A)\exp(\beta_{a0}+\beta_aa)}
\end{equation}
with $\beta_a=\frac{\mu_{A1}-\mu_{A2}}{\sigma^2_{A}}$ and  $\beta_{a0}=\frac{\mu^2_{A2}-\mu^2_{A1}}{2\sigma^2_{A}}$. $\epsilon_A$ is the response bias in the auditory-only trials.

For this model (as well all other models), we fixed the values of the means to be the end-points of the corresponding continuum: $\mu_{A1}=0$ and $\mu_{A2}=4$ (and similarly $\mu_{V1}=0$, and $\mu_{V2}=4$). To determine the values of the bias and the variance, we fitted a model for each modality, collapsed across participants. For the auditory modality, we got $\epsilon_A=-0.20$ and $\sigma^2_A=2.04$. For the visual modality, we got $\epsilon_V=-0.11$ and $\sigma^2_V=3.34$.  In figure \ref{fig:unimodal} (top) we plotted responses in the unimodal conditions as well as the unimodal fits.

\textit{Bimodal condition:} we fitted a model to human responses in the bimodal condition, collapsed across participants. We got $\epsilon=-0.32$, $\sigma^2_{Ab}=5.00$ and $\sigma^2_{Vb}=7.27$.

\textit{Baseline model:} we derived the predictions of the baseline model by using the values of the variances derived from the unimodal conditions, and the response bias derived from the bimodal condition, and by plugging these values into the expression of the posterior \ref{eq:pred}.

In figure \ref{fig:Exp} (top) we plotted the participants' responses in the bimodal condition (left), as well as the prediction of the baseline (middle) and the bimodal fit (right). Correlation of human data vs. the bimodal baseline and fit are shown in figure \ref{fig:corr} (top).

\subsubsection{Discussion}
Compare human responses (figure \ref{fig:Exp}, top left) to the baseline (figure \ref{fig:Exp}, top middle). Qualitatively speaking, the participants showed many patterns of optimal behaviour (as defined by the baseline). Let's, for instance, explore how behaviour changed when participants moved from the auditory-only case (dotted black line) to the bimodal case (solid colored lines).  We see that, similar to the baseline, higher certainty in the visual modality either towards `different' (points 3 and 4 in the visual continuum), or towards `same' (points 1 and 0), generally drove responses accordingly, i.e., either towards more 'different' (as the curves moved up, compared to the auditory-only curve), or towards more `same' responses (as the curves moved down). Moreover, the amount of change was generally a function of the strength of evidence brought by the visual modality, with, for example, responses on visual point 4 being generally `higher' than responses on visual point 3. Similarly, responses on visual 0 were generally `lower' than responses on visual 1. Similar things can be said about the pattern of change from the visual-only (dotted colored horizontal lines) to the bimodal case: at each point along the auditory scale, we see that a point of a given color was generally below its corresponding horizontal line (i.e., same color) when the evidence brought by the auditory modality pointed more toward `same' (point 0 and 1), and was above its corresponding line when evidence pointed more toward `different' (point 3 and 4 along the auditory scale). These qualitative patterns were quite similar to the baseline model. Indeed, we found that the latter explained almost 90\% of the variance  ($r^2=0.89$).

Though we see a qualitative resemblance between human data and the baseline, there were quantitative differences: predictions of the baseline were either a bit too high, or a bit too low. We can see this phenomenon clearly in the shape of the correlation graph (figure \ref{fig:corr}, top left). It is as though participants were less certain about responses than expected. Formally speaking, this translated as an increase in the value of the variance associated with each modality (compare the variance values in the baseline model to the values derived from the bimodal fit).  Since the variance characterizes the reliability of a given cue, it seems as if the bimodal presentation caused reliance on both modalities to decrease.  This effect can possibly be attributed to the arbitrary nature of the form-meaning mapping. In fact, previous studies suggest that while redundant mapping improves performance (e.g., determining the frequency of a bouncing ball from visual and auditory cues), arbitrary mappings generally tends to hinder performance (see \citeA{robinson2010} for a review).

\textit{Modality preference?} we would like to know if part of the observed discrepancy between the baseline and human responses involved a systematic preference (albeit small) for a particular modality when making decisions in the bimodal condition. This preference, if real, should manifest as a deviation from the decision threshold predicted by the baseline model. The decision threshold is defined as the set of values in the audio-visual space along which the posterior (expression \ref{eq:pred}) is equal to 0.5. As illustrated in \ref{fig:space}, above this line participants are more likely to choose one word category, and below this line, they are more likely to choose the other word category. 
The decision threshold takes the following form:
\begin{equation}
v=-\frac{\sigma^2_V}{\sigma^2_A}a+v_0
\label{eq:threshold}
\end{equation}
If the slope derived from the fit is bigger than the slope of the baseline, this would suggest a general preference for the auditory modality, and a smaller slope would suggest a general preference for the visual modality. The limit cases are when there is exclusive reliance on the auditory cue (a vertical line), and where there is exclusive reliance on the visual (a horizontal line). In figure \ref{fig:pref} (top left) we plotted some decision thresholds in the audio-visual space when the intercept was kept constant. The fit to human data  (solid black line) was very close to the baseline threshold (red line). Indeed non-parametric bootstrap sampling over the data showed no proof of significant deviation from the baseline (\ref{fig:pref}, bottom left).

Finally, we found negative values in all response bias terms, which suggests a general bias toward answering `different'.  This bias is probably due to the categorical nature of the task: when two items are ambiguous but perceptually different, this could cause a slight preference for `different' over `same'.

\subsection{experiment 2}

>>>>

\title{Word Identification Under Multimodal Uncertainty}

@@

Identifying the visual referent of a spoken word -- that a particular insect is referred to by the word ``bee'' -- requires both the ability to process and integrate multi-modal input and the ability to identify categories under uncertainty. How do these tasks interact with one another? We introduce a task that allows us to examine how adults identify word categories under joint uncertainty in the auditory and visual modalities. We propose a Bayesian model of the task which provides an optimal baseline. Model predictions are tested in two experiments where category recognition is made under two kinds of uncertainty: ambiguity and noise. In both cases, the optimal baseline model explains much of the variance in human judgments. But when one modality had noise added to it, human perceivers systematically preferred the unperturbed modality to a greater extent than the optimal baseline model did.

@@

 % (e.g., the visual modality in the case of concrete objects)
Language uses symbols expressed in one modality (e.g., the auditory modality, in the case of speech) to communicate about the world, which we perceive through many different sensory modalities. Consider hearing someone yell ``bee!" at a picnic, as a honeybee buzzes around the food. Determining reference involves processing the auditory information and linking it with other perceptual signals (e.g., the visual image of the bee, the sound of its wings, the sensation of the bee flying by your arm). But processing this linguistic input also requires categorization. On the linguistic side,  individual sounds must be identified as members of phonetic categories for the word to be understood. On the referential side, individual sensory impressions must be categorized as coming from a bee. Once these two sets of category-identification tasks have been completed, then a listener can make the referential link between word and world.

This comprehension task takes place in a noisy world. On the auditory side, individual acoustic word tokens are almost always ambiguous with respect to the particular sequence of phonemes they represent \cite<e.g.,>{hillenbrand1995}. And some tokens may be distorted additionally by mispronunciation or ambient noise. Perhaps the speaker was yelling ``pea" and not ``bee.'' Similarly, a sensory impression may not be enough to make a definitive identification of a visual category.\footnote{In the general case, language can of course be visual as well as auditory, and object identification can be done through many modalities. For simplicity, we focus on audio-visual matching here.} Perhaps the quickly flying insect was a beetle or a fly instead.

Thus, establishing reference requires reasoning under a great deal of uncertainty. Work on phoneme categorization has suggested that perceivers encode tokens probabilistically: they base their decisions on these probabilities and adjust their behavior when these probabilities change \cite<e.g.,>{clayard08}. However, previous work in this line of research has largely focused on the unimodal case, that is, when category uncertainty is only in one modality \cite<e.g.,>{feldman2009}. A few studies have explored some aspects of cross-modal processing  \cite<e.g.,>{bejjanki2011, kleinschmidt2015}. This work has focused on the specific case of phoneme recognition from speech and lip movement, however, where information from both modalities is tightly correlated.

In the present work, we study the case of word recognition, where a word is an \textit{arbitrary} mapping between a sound category and a visual category. We test participants on their ability to process audio-visual stimuli and use them to recognize the underlying word category. Crucially, this recognition is made under uncertainty about both the auditory category (a single phoneme, in our case) and the visual category (the concept being referred to). Our goal is to explore how adults deal with such multimodal ambiguity in a categorization task. To this end, we propose a model that combines uncertainty in an optimal way, and we use it as a baseline for interpreting human data. More precisely, we examine the extent to which human data conform to, or deviate from the baseline, and whether the deviation involves a systematic preference for a particular modality.

The paper is organized as follows. First, we introduce our audio-visual word recognition task. We next present the model and explain how it provides an optimal baseline. Then we present two behavioral experiments where we test category recognition under audio-visual uncertainty. In Experiment 1, audio-visual tokens are clear but their category membership is ambiguous. In Experiment 2, we intervene by adding additional perceptual noise to one modality. While participants show no modality preference in Experiment 1, in Experiment 2 they over-rely on visual input when the auditory modality is noisy.

\section{The Audio-Visual Word Recognition Task}

@@

In each trial, participants are presented with an audio-visual target (the prototype of the target category), immediately followed by an audio-visual test stimulus (Figure \ref{fig:task}). The test stimulus may differ from the target in both the auditory and the visual components. The auditory component is sampled from a continuum along the second formant (F2) linking the words /aba/ and /ada/. The visual component is sampled from a morph that links the prototypical pictures of the dog and the cat.  After these two presentations, participants are instructed to press ``same'' or ``different.''

This task is similar to the task introduced by \citeA{sloutsky2003} and used in subsequent research to probe audio-visual encoding. However, unlike this previous line of research, here participants are not asked whether the two audio-visual presentations are identical. Instead, the task is category-based: They are asked to press `same' if they think the second item (the test) belonged to the same category as the first (e.g.,  dog-/aba/), even if there is a slight difference in the word, in the object, or in both. They are instructed to press `different' only if they think that the second stimulus was an instance of the other word category (cat-/ada/).

The task also included trials where pictures were hidden (audio-only) or where sounds were muted (visual-only). These unimodal trials provide us with participants' categorization functions for the auditory and visual categories and are used as inputs to the optimal model, described below.

@@

\section{Ideal Observer Model}

The basis of our ideal observer model is that individual categorization functions from each modality should be combined optimally.
In each modality, we have two categories: /ada/ ($A=1$) and /aba/ ($A=2$) in the auditory dimension, and \textit{cat} ($V=1$) and \textit{dog} ($V=2$) in the visual dimension.
% \footnote{The identity of the categories was randomized in the experimental setup.}
We assume, for the sake of simplicity, that the probability of membership in each category is normally distributed:

$$ p(a | A) \sim  N(\mu_A, \sigma^2_A) $$
$$ p(v | V) \sim  N(\mu_V, \sigma^2_V) $$

In the bimodal condition, participants see word tokens with audio-visual input, and have to make a categorization decision. We represent word tokens as vectors in the audio-visual space, $\mathbf{w}=(a,v)$.
A word category $W$ is defined as the joint distribution of auditory and visual categories. It can be characterized with a bivariate normal distribution in the audio-visual space:
$$ p(\mathbf{w} | W) \sim  N(M_W, \Sigma_W) $$

\begin{figure}[tp]
  \centering
  \includegraphics[width=3.25in]{MyTask.png}
  \caption{Illustration of our model using simulated data. A word category is defined as the joint bivariate distribution of an auditory category (horizontal, bottom panel) and a visual semantic category (vertical, left panel). Upon the presentation of a word token $w$, participants guess whether it is sampled from the word category $W_1$ or from $W_2$. Decision threshold is where the guessing probability is 0.5.}
  \label{fig:space}
\end{figure}

We have two word categories: dog-/aba/ ($W_1$) and cat-/ada/ ($W_2$). Participants can be understood as choosing one of these two word categories (figure \ref{fig:space}). For an ideal ``recognizer'', the probability of choosing category 2 when presented with an audio-visual instance $\mathbf{w}=(a,v)$ is the posterior probability of this category:

\begin{equation}
% \small
p(W_2 | \mathbf{w})=\frac{p(\mathbf{w}|W_2)p(W_2)}{p(\mathbf{w}|W_2)p(W_2)+p(\mathbf{w}|W_1)p(W_1)}
\end{equation}

We make the assumption that, given a particular word category, the auditory and visual tokens are independently generated:

\begin{equation}
p(\mathbf{w} | W) = p(a,v| W) = p(a| W)p(v| W)
\end{equation}

Under this assumption, the posterior probability reduces to:
\begin{equation}
p(W_2 | \mathbf{w})=\frac{1}{1+(1+\epsilon)\exp(\beta_0+\beta_aa+\beta_vv)}
\label{eq:pred}
\end{equation}

@@

and $1+\epsilon=\frac{p(W_1)}{p(W_2)}$ is the proportion of the prior probabilities. If the identity of word categories is randomized, and if $W_1$ is the target, then $\epsilon$ measures a response bias to ``same'' if $\epsilon > 0 $, and a response bias to ``different'' if $\epsilon < 0 $.

@@

\section{Experiment 1}

In experiment 1, we test the predictions of the model in the case where uncertainty is due to perceptual ambiguity in terms of category membership. This experiment had two goals. First, we wanted to examine the extent to which human responses corresponded to the predictions of the ideal observer model. Second, if there was a deviation from the baseline, we wanted to be able to measure whether this deviation involved a systematic preference for a particular modality.

\subsection{Methods}

\subsubsection{Participants}

We recruited a planned sample of 100 participants, recruited from Amazon Mechanical Turk. Only participants with US IP addresses and a task approval rate above 85\% were allowed to participate. They were paid at an hourly rate of \$6/hour. Data were excluded if participants completed the task more than
once (2 participants). Moreover, as specified in the preregistration (\url{https://osf.io/h7mzp/}), participants were excluded if they had less than 50\% accurate responses on the unambiguous training trials (6), and if they reported having experienced a technical problem of any sort during the online experiment (14). The final sample consisted of 76 participants.

\subsubsection{Stimuli}

For  auditory stimuli, we used the continuum introduced in \citeA{vroomen2004}, a 9-point /aba/--/ada/ speech continuum created by varying the frequency of the second (F2) formant in equal steps. We selected 5 equally spaced points from the original continuum by keeping the end-points (prototypes) 1 and 9, as well as points 3, 5, and 7 along the continuum. For visual stimuli, we used a morph continuum introduced in \citeA{freedman2001}: a continuous set of images was generated from three cat prototypes and three dog prototypes. From the original 14 points, we selected 5 points as follows: we kept the item that seemed most ambiguous (point 8), the 2 preceding points (i.e., 7 and 6) and the 2 following points (i.e., 8 and 9). The 6 and 9 points along the morph were quite distinguishable, and we took them to be our prototypes.

 % (Figure \ref{fig:morph})
%
% \begin{figure}[tp]
% \centering
% \includegraphics[width=0.4\textwidth]{morph.png}
% \caption{Visual stimuli for our experiments.}
% \label{fig:morph}
% \end{figure}

\subsubsection{Design and Procedure}

We told participants that an alien was naming two objects: a dog, called /aba/ in the alien language, and a cat, called /ada/. In each trial, we presented the first object (the target) on the left side of the screen simultaneously with the corresponding sound. The target was always the same (e.g., dog-/aba/). The second sound-object pair (the test) followed on the other side of the screen after 500ms and varied in its category membership. For both the target and the test, visual stimuli were present for the duration of hte sound clip ($\sim$800ms). We instructed participants to press ``S'' for same if they thought the alien was naming another dog-/aba/, and ``D'' for different if they thought the alien was naming a cat-/ada/. For each participant, we randomized the sound-object mapping as well as the identity of the target.

The first part of the experiment trained participants using only the prototype pictures and the prototype sounds (12 trials, 4 each from the bimodal, audio-only, and visual-only conditions). After completing training, we instructed participants on the structure of the task and encouraged them to base their answers on both the sounds and the pictures (in the bimodal condition). There were a total of 25 possible combinations in the bimodal condition, and 5 in each of the unimodal conditions. Each participant saw each possible trial in each condition twice, for a total of 70 trials/participants. Trials were blocked by condition and blocks were presented in random order.
% We designed the experiment to take between 10 to 15 minutes to complete.

\begin{figure}
\centering
\includegraphics[width=3.25in\textwidth]{unimodal.png}
\caption{Average human responses in the auditory-only condition (left), and visual-only condition (right). Error bars are 95\% confidence intervals. Solid lines represent unimodal logistic fits.}
\label{fig:unimodal}
\end{figure}

\subsection{Results}
%
% \begin{figure}[h]
% \centering
% \includegraphics[width=3.25in\textwidth]{correlation.png}
% \caption{Human responses vs. the baseline predictions. Error bars are 95\% confidence intervals. In experiment 1, the baseline and the fit explained 89\%, and 94\% of total variance, respectively. In experiment 2, the baseline and the fit explained 91\%, and 97\% of total variance, respectively}
% \label{fig:corr}
% \end{figure}

\begin{figure*}[t]
\centering
\includegraphics[width=.83\textwidth]{Exp.png}
\caption{Proportion of ``different'' judgments as a function of auditory distance. Solid lines represent average human responses (left), predictions of the baseline (middle), and the bimodal fit (left). Dashed lines represent average human responses in the unimodal conditions. Colors represent values in the visual continuum.}
\label{fig:Exp}
% \vspace{10ex}
\end{figure*}

% Let's take the auditory-only case as an example,
\subsubsection{Unimodal conditions}

Average categorization judgments and fits are shown in Figure \ref{fig:unimodal}. Auditory judgements were somewhat sharper and more pronounced than visual judgments. For an ideal recognizer, the probability of choosing category 2 (that is, to answer ``different'') when presented with e.g., an audio instance $a$, is the posterior probability of this category $p(A_2|a)$. If we assume that both categories have equal variances, the posterior probability reduces to:

\begin{equation}
p(A_2 | a)=\frac{1}{1+(1+\epsilon_A)\exp(\beta_{a0}+\beta_aa)}
\end{equation}

with $\beta_a=\frac{\mu_{A_1}-\mu_{A_2}}{\sigma^2_{A}}$ and  $\beta_{a0}=\frac{\mu^2_{A_2}-\mu^2_{A_1}}{2\sigma^2_{A}}$. $\epsilon_A$ is the response bias in the auditory-only trials.

For this model (as well all other models), we fixed the values of the means to be the end-points of the corresponding continuum: $\mu_{A1}=0$ and $\mu_{A2}=4$ (and similarly $\mu_{V1}=0$, and $\mu_{V2}=4$). To determine the values of the bias and the variance, we fit a model for each modality, collapsed across participants. For the auditory modality, we got $\epsilon_A=-0.20$ and $\sigma^2_A=2.04$. For the visual modality, we got $\epsilon_V=-0.11$ and $\sigma^2_V=3.34$.
% In figure \ref{fig:unimodal} (top) we plotted responses in the unimodal conditions as well as the unimodal fits.

\subsubsection{Bimodal condition}

We fit a model to human responses in the bimodal condition, collapsed across participants, finding $\epsilon=-0.32$, $\sigma^2_{Ab}=5.00$ and $\sigma^2_{Vb}=7.27$.

\subsubsection{Baseline model}

We derived the predictions of the baseline model by using the values of the variances derived from the unimodal conditions, and the response bias derived from the bimodal condition, and by substituting these values into the expression of the posterior in Eq.~\ref{eq:pred}. Figure \ref{fig:Exp} shows participants' responses in the bimodal condition (left), as well as the prediction of the baseline (middle), and the bimodal fit models (right).
% Correlation of human data and the bimodal baseline/fitted model are shown in Figure \ref{fig:corr} (top).

\subsubsection{Response bias}

We found negative values for in all response bias terms, which suggests a general bias toward answering ``different.''  This bias is probably due to the categorical nature of the task: when two items are ambiguous but perceptually different, this could cause a slight preference for ``different'' over ``same.''

\subsubsection{Modality preferences}

We next analyzed whether there was a preference for one or the other modality when making decisions in the bimodal condition, beyond that explained by the variance in categories implied by the unimodal responses. This preference would manifest as a deviation from the decision threshold predicted by the baseline model. The decision threshold is defined as the set of values in the audio-visual space along which the posterior (Eq.~\ref{eq:pred}) is equal to 0.5. The decision threshold takes the following form:
% As illustrated in \ref{fig:space}, above this line participants are more likely to choose one word category, and below this line, they are more likely to choose the other word category.

\begin{equation}
v=-\frac{\sigma^2_V}{\sigma^2_A}a+v_0
\label{eq:threshold}
\end{equation}

If the slope derived from the fit is bigger than the slope of the baseline, this finding would suggest a general preference for the auditory modality (similarly, a smaller slope would suggest a preference for the visual modality). The limit cases are when there is exclusive reliance on the auditory cue (a vertical line), and where there is exclusive reliance on the visual (a horizontal line). Figure \ref{fig:pref} (top left) shows the decision threshold in the audio-visual space with a constant intercept; the fit to human data  (solid black line) was very close to the baseline threshold (red line). Non-parametric resampling sampling of the data showed no evidence of a deviation from the baseline (\ref{fig:pref}, bottom left).

\subsection{Discussion}

% Compare human responses (figure \ref{fig:Exp}, top left) to the baseline (figure \ref{fig:Exp}, top middle).
Qualitatively, participants' judgments were similar to the predictions of the optimal baseline model (when fit to the unimodal data). Consider the contrast between the auditory-only case (dashed black line in Figure \ref{fig:Exp}, top left) and the bimodal case (solid colored lines).  Higher certainty in the visual modality generally influenced responses, with greater visual distance leading to more ``different'' ratings and less visual distance leading to more ``same'' ratings. Indeed, we found that the baseline model explained almost 90\% of the variance in judgments ($r^2=0.89$). In addition, as described above, we did not see any evidence for a modality preference.

But although we see a qualitative resemblance data and baseline model, there were quantitative differences. Model predictions were more influenced by the visual modality at the auditory midpoint (the point of highest uncertainty) than human judgements, and were more compressed at the endpoints (the points of lowest uncertainty). Formally speaking, this trend translated into an increase in the value of the variance associated with each modality (e.g., compare the variance values in the baseline model to the values derived from the bimodal fit).  Since the variance characterizes the reliability of a given cue, it seems as if bimodal presentation caused reliance on both modalities to decrease. This effect could possibly be attributed to the arbitrary nature of the form-meaning mapping. In fact, previous studies suggest that while redundant multi-modal information improves performance (e.g., determining the frequency of a bouncing ball from visual and auditory cues), arbitrary mappings generally tends to hinder performance \citeA<for review, see>{robinson2010}.

\section{Experiment 2}

In Experiment 1, we tested the ability of people to recognize a word category under ambiguous tokens. In particular, the tokens were clear but their category membership was uncertain. In real life situations, however, even the identity of the token can be uncertain due to various distorting factors. In order to account for this kind of uncertainty in the auditory modality, we can imagine \cite<following>{feldman2009}, that the speaker generates a target production $t$ from an auditory category
$t | A \sim N(\mu_{A}, \sigma^2_{A})$. The listener, in a noisy condition, does not hear exactly this produced target, but the target perturbed by normally distributed noise: $a | t \sim N(t, \sigma^2_{N})$. When we integrate over $t$, we get:

\begin{equation}
a | A \sim N(\mu_{A}, \sigma^2_{A}+\sigma^2_{N})
\end{equation}
